Lensing in the Schwarzschild Spacetime Applicable to Neutron Stars

### 2023-08-20 src/tex/ms.tex
computing E

Let's calculate $\frac{d\chi^t(\lambda)}{d\lambda}$ and $\frac{d\chi^r(\lambda)}{d\lambda}$ assuming $\theta=\pi/2$ and therefore $p^\theta = 0$. The we have
in null geodesic:
%
\begin{linenomath}\begin{align}
    \label{eq:nullgeo}
    -(1-\frac{2}{r})(p^t)^2 + \frac{1}{1-\frac{2}{r}}(p^r)^2 + r^2\sin^2 \theta (p^\phi)^2 = 0
\end{align}\end{linenomath}
%
Solving eq. \ref{eq:nullgeo} leads to $p^t = \frac{d\chi^t(\lambda)}{d\lambda} = \frac{e}{1-\frac{2}{r}}$ and 
%
\begin{linenomath}\begin{align}
    \label{eq:pr-squared}
    \left(\frac{d\chi^r(\lambda)}{d\lambda}\right)^2 = \left(\frac{e^2}{1-\frac{2}{r}} - \frac{l^2}{r^2}\right) \left(1-\frac{2}{r}\right)
\end{align}\end{linenomath}
%
which, in terms of $b$ becomes
%
\begin{linenomath}\begin{align}
    \label{eq:pr-b}
    \frac{d\chi^r(\lambda)}{d\lambda} = e\sqrt{\left(\frac{1}{1-\frac{2}{r}} - \frac{b^2}{r^2}\right) \left(1-\frac{2}{r}\right)}
\end{align}\end{linenomath}
%
Therefore, the integral \ref{eq:timedelay} becomes:
%
\begin{linenomath}\begin{align}
    \label{eq:timedelayfull}
    \Delta t = \int_{r_{NS}}^{\infty} dr \left[\left(\frac{1}{1-\frac{2}{r}}\right) - \frac{1}{\sqrt{\left(1-\frac{2}{r}\right)^2 - \frac{b^2}{r^2}\left(1-\frac{2}{r}\right)^3}}\right]
\end{align}\end{linenomath}
%
Following the substitution in Eq. \ref{eq:w-substitution}, we then get
%
\begin{linenomath}\begin{align}
    \label{eq:tdelay-with-w}
    \Delta t = r_{NS} \int_{1}^{\infty} \mathrm{d} w \, \left[\left(\frac{1}{1-\frac{2}{wr_{NS}}}\right) - \frac{1}{\sqrt{\left(1-\frac{2}{wr_{NS}}\right)^2 - \frac{b^2}{(wr_{NS})^2}\left(1-\frac{2}{wr_{NS}}\right)^3}}\right]
\end{align}\end{linenomath}
%
It is also convenient to make the substitution 
\begin{equation}
    b = x \bNS = x \rNS \left( 1 + \frac{1}{\rNS} + \frac{3}{2 \rNS^2} + \ldots \right),
\end{equation}
which produces 
%
\begin{linenomath}\begin{align}
    \label{eq:tdelay-with-x}
    \Delta t = r_{NS} \int_{1}^{\infty} \mathrm{d}w \, \left[\left(\frac{1}{1-\frac{2}{wr_{NS}}}\right) - \frac{1}{\sqrt{\left(1-\frac{2}{wr_{NS}}\right)^2 - \frac{x^2 \rNS^2 \left( 1 + \frac{1}{\rNS} + \frac{3}{2 \rNS^2} + \ldots \right)^2}{(wr_{NS})^2}\left(1-\frac{2}{wr_{NS}}\right)^3}}\right]
\end{align}\end{linenomath}
%
Substituting $1/r_{NS} = y$, and expanding around $y=0$, the integral gives
%
\begin{multline}
    \label{eq:tdelay-solved}
    \Delta t = r_{NS} \left( -1 + \sqrt{1 - x^2} \right) \\ \times \left( 1 + \frac{1}{r_{NS}} \left( 1  + \frac{2}{-1 + \sqrt{1 - x^2}} \log\left( \frac{x^2}{2 - 2 \sqrt{1 - x^2}} \right) \right) + \mathcal{O}\left( \frac{1}{r_{NS}^2} \right) \right)
\end{multline}
where $-1<x<1$.
\bibliography{bib}

>>>>

Let's calculate $\frac{d\chi^t(\lambda)}{d\lambda}$ and $\frac{d\chi^r(\lambda)}{d\lambda}$ assuming $\theta=\pi/2$ and therefore $p^\theta = 0$. The we have
in null geodesic:
%
\begin{linenomath}\begin{align}
    \label{eq:nullgeo}
    -(1-\frac{2}{r})(p^t)^2 + \frac{1}{1-\frac{2}{r}}(p^r)^2 + r^2\sin^2 \theta (p^\phi)^2 = 0
\end{align}\end{linenomath}
%
Solving eq. \ref{eq:nullgeo} leads to $p^t = \frac{d\chi^t(\lambda)}{d\lambda} = \frac{e}{1-\frac{2}{r}}$ and 
%
\begin{linenomath}\begin{align}
    \label{eq:pr-squared}
    \left(\frac{d\chi^r(\lambda)}{d\lambda}\right)^2 = \left(\frac{e^2}{1-\frac{2}{r}} - \frac{l^2}{r^2}\right) \left(1-\frac{2}{r}\right)
\end{align}\end{linenomath}
%
which, in terms of $b$ becomes
%
\begin{linenomath}\begin{align}
    \label{eq:pr-b}
    \frac{d\chi^r(\lambda)}{d\lambda} = e\sqrt{\left(\frac{1}{1-\frac{2}{r}} - \frac{b^2}{r^2}\right) \left(1-\frac{2}{r}\right)}
\end{align}\end{linenomath}
%
Therefore, the integral \ref{eq:timedelay} becomes:
%
\begin{linenomath}\begin{align}
    \label{eq:timedelayfull}
    \Delta t = \int_{r_{NS}}^{\infty} dr \left[\left(\frac{1}{1-\frac{2}{r}}\right) - \frac{1}{\sqrt{\left(1-\frac{2}{r}\right)^2 - \frac{b^2}{r^2}\left(1-\frac{2}{r}\right)^3}}\right]
\end{align}\end{linenomath}
%
Following the substitution in Eq. \ref{eq:w-substitution}, we then get
%
\begin{linenomath}\begin{align}
    \label{eq:tdelay-with-w}
    \Delta t = r_{NS} \int_{1}^{\infty} \mathrm{d} w \, \left[\left(\frac{1}{1-\frac{2}{wr_{NS}}}\right) - \frac{1}{\sqrt{\left(1-\frac{2}{wr_{NS}}\right)^2 - \frac{b^2}{(wr_{NS})^2}\left(1-\frac{2}{wr_{NS}}\right)^3}}\right]
\end{align}\end{linenomath}
%
It is also convenient to make the substitution 
\begin{equation}
    b = x \bNS = x \rNS \left( 1 + \frac{1}{\rNS} + \frac{3}{2 \rNS^2} + \ldots \right),
\end{equation}
which produces 
%
\begin{linenomath}\begin{align}
    \label{eq:tdelay-with-x}
    \Delta t = r_{NS} \int_{1}^{\infty} \mathrm{d}w \, \left[\left(\frac{1}{1-\frac{2}{wr_{NS}}}\right) - \frac{1}{\sqrt{\left(1-\frac{2}{wr_{NS}}\right)^2 - \frac{x^2 \rNS^2 \left( 1 + \frac{1}{\rNS} + \frac{3}{2 \rNS^2} + \ldots \right)^2}{(wr_{NS})^2}\left(1-\frac{2}{wr_{NS}}\right)^3}}\right]
\end{align}\end{linenomath}
%
Substituting $1/r_{NS} = y$, and expanding around $y=0$, the integral gives
%
\begin{multline}
    \label{eq:tdelay-solved}
    \Delta t = r_{NS} \left( -1 + \sqrt{1 - x^2} \right) \\ \times \left( 1 + \frac{1}{r_{NS}} \left( 1  + \frac{2}{-1 + \sqrt{1 - x^2}} \log\left( \frac{x^2}{2 - 2 \sqrt{1 - x^2}} \right) \right) + \mathcal{O}\left( \frac{1}{r_{NS}^2} \right) \right)
\end{multline}
where $-1<x<1$.

\section{Creating matrix E for the energy factor}
To express the expression in \ref{eq:energy-relation} in matrix form using the spherical harmonic basis, we need to define matrices 
that correspond to the spherical harmonic functions and their associated coefficients.

Let's denote the spherical harmonic basis as $Y_{lm}(\theta, \phi)$. The spherical harmonic basis is expressed as:
\begin{equation}
    \label{eq:spherical-harmonics}
    Y_{lm}(\theta, \phi) = P_{lm}(\theta, \phi) \cdot e^{i \cdot m \cdot \phi}
\end{equation}

where $P_{lm}(\theta, \phi)$ are the associated Legendre polynomials.

To express the given expression in matrix form using the spherical harmonic basis, we need to expand the function \( F(y, v_{\text{rot}}, \theta_0, r_s) \) 
into a series of spherical harmonic functions and their coefficients. The general form of the spherical harmonic expansion is:

\begin{equation}
    F(y, v_{rot}, \theta_0, r_s) = \sum_{l=0}^{\infty} \sum_{m=-l}^{l} C_{lm} Y_{lm}(\theta, \phi) 
\end{equation}

% Now, let's break down the given function step by step:

% \begin{multline}
%     F(y, v_{rot}, \theta_0, r_s) &= 1 - yv_{\text{rot}}\sin{\theta_0} + (y^2 - \frac{1}{2})v_{\text{rot}}^2\sin^2{\theta_0} - \frac{1}{r_s}\left(1-(y^2 - \frac{1}{2})v_{\text{rot}}^2\sin^2{\theta_0}\right) \\
%     = 1 - yv_{\text{rot}}\sin{\theta_0} + (y^2 - \frac{1}{2})v_{\text{rot}}^2\sin^2{\theta_0} - \\ \frac{1}{r_s} + \frac{(y^2 - \frac{1}{2})v_{\text{rot}}^2\sin^2{\theta_0}}{r_s} Y_{lm}(\theta, \phi) = P_{lm}(\theta, \phi) \cdot e^{i \cdot m \cdot \phi}
% \end{multline}

Now, the spherical harmonic expansion involves integrating \( F \) against each spherical harmonic function \( Y_{lm} \). 
The coefficients \( C_{lm} \) can be found using the following formula:

\begin{linenomath}\begin{align}
    C_{lm} = \int F(y, v_{rot}, \theta_0, r_s) \cdot Y_{lm}^*(\theta, \phi) \, d\Omega 
\end{align}\end{linenomath}

Here's how the matrix elements are formed:

\begin{linenomath}\begin{align}
C_{lm} = \int F(y, v_{\text{rot}}, \theta_0, r_s) \cdot Y_{lm}^*(\theta, \phi) \, d\Omega 
\end{align}\end{linenomath}

For \( C_{00} \) (the \( l = 0, m = 0 \) component):

\begin{linenomath}\begin{align}
    C_{00} = \int F(y, v_{\text{rot}}, \theta_0, r_s) \cdot Y_{00}^*(\theta, \phi) \, d\Omega = \int F(y, v_{\text{rot}}, \theta_0, r_s) \, d\Omega 
\end{align}\end{linenomath}

In the spherical harmonic basis, we can express the coefficients of the basis functions in matrix form:

\begin{multline}
    \mathbf{C} = \begin{bmatrix}
    1 & -v_{\text{rot}}\sin{\theta_0} & v_{\text{rot}}^2\sin^2{\theta_0} - \frac{1}{r_s} & 0 & \ldots \\
    0 & y\sqrt{2}v_{\text{rot}}\sin{\theta_0} & 0 & \ldots & \ldots \\
    0 & 0 & (y^2 - \frac{1}{2})\sqrt{6}v_{\text{rot}}^2\sin^2{\theta_0} & \ldots & \ldots \\
    0 & 0 & 0 & \ldots & \ldots \\
    \ldots & \ldots & \ldots & \ldots & \ldots \\
    \end{bmatrix}
\end{multline}

Here, the rows of the matrix correspond to different values of "l" and the columns correspond to different values of "m". 
The diagonal elements of the matrix correspond to the coefficients of the spherical harmonic basis functions with the same "l" and "m" values, 
while off-diagonal elements are zero.

Spherical harmonics for \(m = -1\), \(m = 0\), and \(m = 1\):
$$Y_{l,-1}(\theta, \phi) = \sqrt{\frac{3}{8\pi}}e^{-i\phi}\sin\theta$$
$$Y_{l,0}(\theta, \phi) = \sqrt{\frac{3}{4\pi}}\cos\theta$$
$$Y_{l,1}(\theta, \phi) = -\sqrt{\frac{3}{8\pi}}e^{i\phi}\sin\theta$$

Matrix elements for \(m = -1\):
$$M_{-1} = -\sqrt{\frac{3}{8\pi}}v_{\text{rot}}e^{-i\phi_0}\sin\theta_0$$

Matrix elements for \(m = 0\):
$$M_{0} = -\frac{\sqrt{6}}{2}v_{\text{rot}}^2e^{-2i\phi_0}\sin^2\theta_0$$

Matrix elements for \(m = 1\):
$$M_{1} = -\sqrt{\frac{3}{8\pi}}v_{\text{rot}}^2e^{-2i\phi_0}\sin^2\theta_0$$

\begin{multline}
    \begin{bmatrix}
        1 & -yv_{\text{rot}}\sin\theta_0 & (y^2 - \frac{1}{2})v_{\text{rot}}^2\sin^2\theta_0 - \frac{1}{r_s}(1 - (y^2 - \frac{1}{2})v_{\text{rot}}^2\sin^2\theta_0) \\
        0 & -\sqrt{\frac{3}{8\pi}}v_{\text{rot}}e^{-i\phi_0}\sin\theta_0 & -\frac{\sqrt{6}}{2}v_{\text{rot}}^2e^{-2i\phi_0}\sin^2\theta_0 \\
        0 & 0 & -\sqrt{\frac{15}{8\pi}}v_{\text{rot}}^2e^{-2i\phi_0}\sin^2\theta_0
    \end{bmatrix}
\end{multline}

\bibliography{bib}
